---
title: 'Ensemble Forecasts and Uncertainty Partitioning'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ecoforecastR)
library(dplyr)
library(lubridate)
library(tidyr)
```

## Load in model run from milestone 5

```{r}
load(file = "DLM.Rdata")
```

### If you don't have the DLM saved, set eval=TRUE to run the chunk to get the DLM

```{r, eval=FALSE}
URL = "https://data.ecoforecast.org/neon4cast-targets/phenology/phenology-targets.csv.gz"

# Get gcc data
data = readr::read_csv(URL, col_names = c("Date", "Siteid", "Variable", "Observation"))
harv = na.omit(data[data$Siteid == "HARV",]) # filter by Harvard Forest site
gcc_90 = harv[harv$Variable == "gcc_90",]

# Get historic temperature data
weather_stage3 <- neon4cast::noaa_stage3()
ds1 <- weather_stage3 |> 
  dplyr::filter(site_id == "HARV") |>
  dplyr::collect()
TMP = ds1[ds1$variable == "air_temperature",]
tmps <- as.data.frame(TMP) # save temperature data as dataframe
tmps$date <- date(tmps$datetime)
tmps$date <- as.character(tmps$date)
# Get maximum temperature within each ensemble per day: 31 ensembles x 36 days
temp_max <- tmps %>% 
  group_by(date) %>%
  summarize(temp.max = max(prediction)) 
data1 <- gcc_90[which(gcc_90$Date == temp_max$date[1]):nrow(gcc_90), ]
temp <- temp_max[temp_max$date %in% data1$Date, ]
# Add temperature as column in data dataframe
data1$Temperature <- temp$temp.max - 273.15
mean_temp <- tmps %>%
  group_by(date) %>%
  summarize(temp.mean = mean(prediction) - 273.15)
mean_temp <- mean_temp[mean_temp$date %in% data1$Date,]
data1$TempDiff <- mean_temp$temp.mean - data1$Temperature

#data_36 <- tail(data1, n=36) # get the last 36 days for prediction?

# Run the DLM
gcc.out <- ecoforecastR::fit_dlm(model=list(obs="Observation", fixed="~ 1 + X + TempDiff"), data1)

# comment out to save DLM as Rdata, so you don't have to run this again
#save(gcc.out, file="DLM.Rdata")
```

## Step 1: Grabbing the Forecast Temperature Data
```{r}
weather_stage3 <- neon4cast::noaa_stage2(start_date = as.character(Sys.Date() - lubridate::days(2)))
ds1 <- weather_stage3 |> 
  dplyr::filter(site_id == "HARV") |>
  dplyr::collect()

TMP = ds1[ds1$variable == "air_temperature",]

TMP <- TMP %>% 
  mutate(datetime = lubridate::as_date(datetime)) %>% 
  group_by(datetime, site_id, parameter, variable) |> 
  summarize(prediction = mean(prediction),.groups = "drop") |> 
  select(datetime, site_id, variable, prediction, parameter)

# clean up temp data
tmps <- as.data.frame(TMP) # save temperature data as dataframe

# Get maximum temperature within each ensemble per day: 31 ensembles x 36 days
temp_max <- tmps %>% 
  group_by(datetime, parameter) %>%
  summarize(temp.max = max(prediction)) 

# columns: ensemble, rows: forecast by time step
temps <- pivot_wider(temp_max, names_from = parameter, values_from = temp.max) 
```


## Step 2: Define a forecast function that takes in the calculated DLM parameters, and uses the forecast temperature

```{r}
##` @param IC    Initial Conditions
##` @param temp  Temperature forecast
##` @param beta  Slope of temperature effect on gcc
##` @param alpha Site random effect
##` @param Q     Process error (default = 0 for deterministic runs)
##` @param n     Size of Monte Carlo ensemble
forecast <- function(IC, temp, beta, Q=0, n, gmin, gmax){ # temp should be a vector
  N <- matrix(NA, NT, n)  ## rows: forecasts by time, columns: ensemble number
  Nprev <- IC           ## initialize
  for(t in 1:NT){ # update daily per ensemble...
    mean_temp <- mean(as.numeric(temp[,-1][t,])) - 273.15
    temp_diff <- mean_temp - (temp[,-1][t,] - 273.15)
    gcc_predict = Nprev + beta * temp_diff # linear model
    N[t,] <- pmax(pmin(rnorm(n, mean(as.numeric(gcc_predict)), Q), gmax), gmin) # Ensure prediction is within bounds and predict next step
    #N[,t] <- rnorm(n, gcc_predict, Q)
    Nprev <- N[t,] # update IC
  }
  return(N)
}
```

## Step 3: Run the forecast on the temperature ensembles and plot the results

```{r}
Nmc = 31 # number of Monte Carlo ensembles
NT = 36 # number of time steps into the future

# extract necessary things from DLM
params <- window(gcc.out$params, start=1000)
out <- as.matrix(gcc.out$predict) # get model predictions
ci <- apply(out, 2, quantile, c(0.025,0.5,0.975)) # get model confidence intervals
data <- gcc.out$data

test <- as.data.frame(params[[1]])
param.mean <- apply(params[[1]], 2, mean)
betaTemp <- param.mean["betaTempDiff"]  # get the mean beta for temperature
q <- 1 / sqrt(param.mean["tau_add"])

prow <- sample.int(nrow(params[[1]]), Nmc, replace=TRUE)
Qmc <- 1 / sqrt(params[[1]][prow, "tau_add"])
drow <- sample.int(ncol(temp_max), Nmc, replace=TRUE)

IC <- rnorm(Nmc, tail(as.numeric(data$OBS,1)), tail(Qmc, 1))

pheno_forecast <- forecast(IC = IC,#as.numeric(tail(gcc.out$predict[[1]], n=1)), IC = mean(as.numeric(data$OBS))
                           temp = temps,
                           beta = betaTemp,
                           Q = Qmc,
                           n = 36,
                           gmin = min(as.numeric(data$OBS), na.rm=T),
                           gmax = max(as.numeric(data$OBS), na.rm=T))
# Plot the ensemble runs
for(ensemble in 1:31) {
  plot(pheno_forecast[,ensemble],
       type='b',
       xlab="time",
       ylab="prediction gcc",
       main = c(ensemble, "ensemble number"))
}
```

# Uncertainty Analysis

## Set up the plot function
```{r}
s <- 1             ## Focal site for forward simulation
#Nmc = 1000         ## set number of Monte Carlo draws
ylim = c(0.3,0.4)  ## set Y range on plot
N.cols <- c("black","red","green","blue","orange") ## set colors
trans <- 0.8       ## set transparancy
time = 1:(NT*2)    ## total time
time1 = 1:NT       ## calibration period
time2 = time1+NT   ## forecast period

plot.run <- function(){
  #sel = seq(s, ncol(ci), by=1)
  plot(time,time,type='n',ylim = ylim, ylab="gcc")
  ecoforecastR::ciEnvelope(time1, tail(ci[1,], n=36), tail(ci[3,], n=36), col=col.alpha("lightBlue",0.6))
  #points(time1, tail(as.numeric(gcc.out$predict[[1]]), n = 36))
  points(time1, tail(as.numeric(gcc.out$predict[[1]]), n = 36))
  lines(time1, tail(ci[2,], n=36), col="blue")
}
```

### Prediction from DLM (first ensemble)
```{r,echo=FALSE}
ci <- apply(tail(gcc.out$predict[[1]], n=36), 2, quantile, c(0.025,0.5,0.975))
plot.run()
lines(time2, pheno_forecast[,1],col="purple",lwd=3)
```

### Initial condition uncertainty
```{r}
#prow = sample.int(nrow(params),Nmc,replace=TRUE)
#IC = as.matrix(gcc.out$predict)
#IC <- rnorm(Nmc, tail(as.numeric(data$OBS,1)), tail(Qmc, 1))

N.I <- forecast(IC=IC, # , IC = mean(as.numeric(data$OBS))
                temp = temps,
                beta = betaTemp,
                Q = Qmc,
                n = 36,
                gmin = min(as.numeric(data$OBS), na.rm=T),
                gmax = max(as.numeric(data$OBS), na.rm=T))

## Plot run
plot.run()
N.I.ci = apply(N.I,2,quantile,c(0.025,0.5,0.975))
ecoforecastR::ciEnvelope(time2, N.I.ci[1,], N.I.ci[3,], col="lightgreen")#col=col.alpha(N.cols[1],trans))
lines(time2,pheno_forecast[1,],lwd=0.5)
```

### Paramter uncertainty (betaTempDiff)
```{r}
N.IP <- forecast(IC=pheno_forecast[1,], # , IC = mean(as.numeric(data$OBS))
                temp = temps,
                beta = params[[1]][prow, "betaTempDiff"],
                Q = Qmc,
                n = 36,
                gmin = min(as.numeric(data$OBS), na.rm=T),
                gmax = max(as.numeric(data$OBS), na.rm=T))

plot.run()
N.IP.ci = apply(N.IP,2,quantile,c(0.025,0.5,0.975))
ecoforecastR::ciEnvelope(time2,N.IP.ci[1,],N.IP.ci[3,],col="pink")#col.alpha(N.cols[2],trans))
ecoforecastR::ciEnvelope(time2, N.I.ci[1,], N.I.ci[3,], col="lightgreen")#col=col.alpha(N.cols[1],trans))
lines(time2,pheno_forecast[1,],lwd=0.5)

```
# Driver uncertainty

```{r}
drow = sample.int(nrow(pheno_forecast),Nmc,replace=TRUE)
N.IPD <- forecast(IC=pheno_forecast[1,], # , IC = mean(as.numeric(data$OBS))
                temp = temps[drow,],
                beta = params[[1]][prow, "betaTempDiff"],
                Q = Qmc,
                n = 36,
                gmin = min(as.numeric(data$OBS), na.rm=T),
                gmax = max(as.numeric(data$OBS), na.rm=T))

plot.run()
N.IPD.ci = apply(N.IPD,2,quantile,c(0.025,0.5,0.975), na.rm=TRUE)
ecoforecastR::ciEnvelope(time2,N.IPD.ci[1,],N.IPD.ci[3,],col=col.alpha(N.cols[3],trans))
ecoforecastR::ciEnvelope(time2,N.IP.ci[1,],N.IP.ci[3,],col=col.alpha(N.cols[2],trans))
ecoforecastR::ciEnvelope(time2, N.I.ci[1,], N.I.ci[3,], col=col.alpha(N.cols[1],trans))
lines(time2,pheno_forecast[1,],lwd=0.5,col="purple")
```

### Process uncertainty : Do not run (its wrong rn)
```{r}
#Qmc <- 1/sqrt(params[prow,"Q"])  ## convert from precision to standard deviation

N.IPDE <- forecast(IC=pheno_forecast[1,], # , IC = mean(as.numeric(data$OBS))
                temp = temps[drow,],
                beta = params[[1]][prow, "betaTempDiff"],
                Q = Qmc,
                n = 36,
                gmin = min(as.numeric(data$OBS), na.rm=T),
                gmax = max(as.numeric(data$OBS), na.rm=T))

## Plot run
plot.run()
N.IPDE.ci = apply(N.IPDE,2,quantile,c(0.025,0.5,0.975), na.rm=TRUE)
ecoforecastR::ciEnvelope(time2,N.IPD.ci[1,],N.IPD.ci[3,],col=col.alpha(N.cols[3],trans))
ecoforecastR::ciEnvelope(time2,N.IP.ci[1,],N.IP.ci[3,],col=col.alpha(N.cols[2],trans))
ecoforecastR::ciEnvelope(time2, N.I.ci[1,], N.I.ci[3,], col=col.alpha(N.cols[1],trans))
ecoforecastR::ciEnvelope(time2,N.I.ci[1,],N.I.ci[3,],col=col.alpha(N.cols[1],trans))
lines(time2,N.I.ci[2,],lwd=0.5)
```
